Efficient simulation of individual-based population models

### 2024-03-27 article.qmd
solve conflict

To summarize, the space of event types is $E = \{ b, en, d, s \}$, and the jump $\Delta Z_t = Z_t - Z_{t^-}$ (change in the population composition) generated by an event of type $e \in \{ b, en, d, s \}$ is denoted by $\phi^e(t, I)$. We thus have the following rules summarized in the table @tbl-event-action.

@@

**Death events** A baseline age and time specific death rate is first
calibrated on "England and Wales (EW)" males mortality historic
data[^1], and projected for 30 years using the Lee-Carter model with the
package `StMoMo` (see [@stmomo]). The forecasted baseline death
intensity is denoted by $d(t,a)$, defined by:
$$
d(t,a) = \sum_{k=0}^{29}\mathbf{1}_{\{k\leq t < k+1\}} d_k(a), \quad \forall \; t\in [0,30] \text{ and } a \in [65, a_{\max}],
$$ {#eq-insurance-baseline}
with $d_k(a)$ the point estimate of the forecasted mortality rate for
age $a$ and year $k$.\
Individuals in risk class 1 are assumed to have mortality rates that are
20% higher than the baseline mortality (for instance, the risk class
could refer to smokers), while individuals in risk class 2 are assumed
to have mortality rates that are 20% lower than the baseline (non
smokers). The death intensity of an individual $I= (\tau_b, \infty, x)$,
of age $a(I,t) = t - \tau_b$ at time $t$ and in risk class
$x \in \{1, 2\}$ is thus the function $$\label{eq-insurance-deathrates}
\lambda^d(t,I) = \alpha_x d(t,a(I,t)), \quad \alpha_1 = 1.2, \quad \alpha_2 = 0.8.$$
In particular, the death intensity verifies Assumption TODO
[2](#AssumptionIntensity1){reference-type="ref"
reference="AssumptionIntensity1"} since:
$$
\lambda^d(t,I) \leq \bar d : = \alpha_1 \sup_{t \in [0,30]} d(t,a_{\max}).
$$ {#eq-insurance-bound-deathrates}

@@

The death and exit intensities are of class `individual` (see Table
@tbl-intensity-classes). Hence, the death and exit events
are created with the `mk_event_individual` function.

@@

Initially in the portfolio (at $t=0$), there is the same number of 65
years old policyholders in each risk class. However, policyholders in
the risk class 2 with lower mortality rates leave the portfolio at
higher rate than policyholders in the risk class 1 : $\mu^2 > \mu^1$.
Therefore, the heterogeneous portfolio composition changes with time,
including more and more individuals in risk class 1 with higher
mortality rates, but with variations across age classes. To illustrate
the composition of the total population at the end of the simulation
($t=30$), we present in Figure @fig-insur-a the age pyramid of the final composition of
the portfolio obtained with the `age_pyramid` and `plot` function of the
`pyramid` class.

@@

In Figure @fig-insur-b, we illustrate the central death rates in the
simulated portfolio at final time. Due to the mortality differential
between risk class 1 and 2, one would expect to observe more individuals
in risk class 2 at higher ages. However, due to exit events, a higher
proportion of individuals in risk class 1 exit the portfolio over time,
resulting in a greater proportion of individuals in risk class 1 at
higher ages than what would be expected in the absence of exit events.
Consequently, the mortality rates in the portfolio are more aligned with
those of risk class 1 at higher ages. This is a simple example of how
composition changes in the portfolio can impact aggregated mortality
rates and potentially compensate or reduce an overall mortality
reduction (see also [@KAAKAI201916]).

@@

![Full vs random algorithm computztion time](img/time_pop_size.png){#fig-fullvsrandom width=80%}

@@

TODO

>>>>

To summarize, the space of event types is $E = \{ b, en, d, s \}$, and the jump $\Delta Z_t = Z_t - Z_{t^-}$ (change in the population composition) generated by an event of type $e \in \{ b, en, d, s \}$ is denoted by $\phi^e(t, I)$. We thus have the following rules summarized in @tbl-event-action.

@@

**Death events** A baseline age and time specific death rate is first
calibrated on "England and Wales (EW)" males mortality historic
data[^1], and projected for 30 years using the Lee-Carter model with the
package `StMoMo` (see [@stmomo]). The forecasted baseline death
intensity is denoted by $d(t,a)$, defined by:
$$
d(t,a) = \sum_{k=0}^{29}\mathbf{1}_{\{k\leq t < k+1\}} d_k(a), \quad \forall \; t\in [0,30] \text{ and } a \in [65, a_{\max}],
$$ {#eq-insurance-baseline}
with $d_k(a)$ the point estimate of the forecasted mortality rate for
age $a$ and year $k$.\
Individuals in risk class 1 are assumed to have mortality rates that are
20% higher than the baseline mortality (for instance, the risk class
could refer to smokers), while individuals in risk class 2 are assumed
to have mortality rates that are 20% lower than the baseline (non
smokers). The death intensity of an individual $I= (\tau_b, \infty, x)$,
of age $a(I,t) = t - \tau_b$ at time $t$ and in risk class
$x \in \{1, 2\}$ is thus the function $$
\lambda^d(t,I) = \alpha_x d(t,a(I,t)), \quad \alpha_1 = 1.2, \quad \alpha_2 = 0.8.
$$
In particular, the death intensity verifies @cnj-intensity2 since:
$$
\lambda^d(t,I) \leq \bar d : = \alpha_1 \sup_{t \in [0,30]} d(t,a_{\max}).
$$ {#eq-insurance-bound-deathrates}

@@

The death and exit intensities are of class `individual` (see
@tbl-intensity-classes). Hence, the death and exit events
are created with the `mk_event_individual` function.

@@

Initially in the portfolio (at $t=0$), there is the same number of 65
years old policyholders in each risk class. However, policyholders in
the risk class 2 with lower mortality rates leave the portfolio at
higher rate than policyholders in the risk class 1 : $\mu^2 > \mu^1$.
Therefore, the heterogeneous portfolio composition changes with time,
including more and more individuals in risk class 1 with higher
mortality rates, but with variations across age classes. To illustrate
the composition of the total population at the end of the simulation
($t=30$), we present in @fig-insur-a the age pyramid of the final composition of
the portfolio obtained with the `age_pyramid` and `plot` function of the
`pyramid` class.

@@

In @fig-insur-b, we illustrate the central death rates in the
simulated portfolio at final time. Due to the mortality differential
between risk class 1 and 2, one would expect to observe more individuals
in risk class 2 at higher ages. However, due to exit events, a higher
proportion of individuals in risk class 1 exit the portfolio over time,
resulting in a greater proportion of individuals in risk class 1 at
higher ages than what would be expected in the absence of exit events.
Consequently, the mortality rates in the portfolio are more aligned with
those of risk class 1 at higher ages. This is a simple example of how
composition changes in the portfolio can impact aggregated mortality
rates and potentially compensate or reduce an overall mortality
reduction (see also [@KAAKAI201916]).

@@

![Full vs random algorithm computation time](img/time_pop_size.png){#fig-fullvsrandom width=80%}

@@

For ease of notation, we prove
the case when $\mathcal{P} =\emptyset$ (there are no events with Poisson
intensity).\
Let $Z$ be the population process obtained by @alg-interaction, and $(T_n)_{n\geq 0}$ the sequence
of its jump times ($T_0=0$).\
**Step 1** Let $T_1$ be the first event time in the population, with its
associated marks defining the type $E_1$ of the event and the individual
$I_1$ to which this event occurs. By construction, $(T_1, E_1, I_1)$ is
characterized by the first jump of:
$$Q^0(\mathrm{d}t, \mathrm{d}k , \mathrm{d}e) = \int_{\mathbb R^+} \mathbf{1}_{\{\theta \leq \lambda_{t}^e(I_k,Z_0)\}}\bar Q^0 (\mathrm{d}t ,\mathrm{d}k , \mathrm{d}e, \mathrm{d}\theta ),$$
with $\bar Q^0$ the Poisson measure introduced in the first step of the
algorithm described in @sec-simulation-algo.

Since $T_1$ is the first event time, the population composition stays
constant, $Z_t=Z_0$, on $\{t<T_1\}$. In addition, recalling that the
first event has the action $\phi^{E_1}(T_1, I_1)$ (see @tbl-event-action) on
the population $Z$, we obtain that: $$\begin{aligned}
Z_{t\wedge T_1} & =  Z_0 + \mathbf{1}_{\{t\geq T_1\}} \phi^{E_1} (T_1 , I_1)  \\
 & = Z_0 + \int_0^{t\wedge T_1}  \int_{\mathcal{J}_0} \phi^e (s , I_k)  Q^0 (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e ) \\
& = Z_0 + \int_0^{t\wedge T_1} \int_{\mathcal{J}_0}  \int_{\mathbb R^+} \phi^e (s , I_k)  \mathbf{1}_{\{\theta \leq \lambda_s^e(I_k,Z_0)\}}\bar Q^0 (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e, \mathrm{d}\theta ).
\end{aligned}$$ Since $Z_{s^-} = Z_0$ on $\{s \leq T_1\}$, the last
equation can be rewritten as $$
 Z_{t\wedge T_1}   = Z_0 + \int_0^{t\wedge T_1} \int_{\mathcal{J}_0}  \int_{\mathbb R^+} \phi^e (s , I_k)  \mathbf{1}_{\{\theta \leq \lambda_s^e(I_k,Z_{s^-})\}}\bar Q^0 (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e, \mathrm{d}\theta ).$$

**Step 2** The population size at the $n$--th event time $T_n$ is
$N_{T_n}$. The $(n+1)$--th event type and the individual to which this
event occur are thus chosen in the set
$$\mathcal{J}_n := \{ 1,\dots, N_{T_n}\} \times (\mathcal{E} \cup \mathcal{E}_W).$$
Conditionally to $\mathcal{F}_{T_n}$, let us first introduce the marked
Poisson measure $\bar Q^n$ on
$[T_n, \infty) \times \mathcal J_n \times \mathbb R^+$, of intensity:
$$
\begin{aligned}
\bar \mu^n(\mathrm{d}t, \mathrm{d}k, \mathrm{d}e , \mathrm{d}\theta ) & := \mathbf{1}_{\{t > T_n \}}\bar \Lambda (N_{T_n})\mathrm{d}t  \frac{\bar \lambda^e }{\bar \Lambda(N_{T_n})} \delta_{\mathcal J_n}(\mathrm{d}k, \mathrm{d}e) \frac{1}{\bar \lambda^e} \mathbf{1}_{[0,\bar \lambda^e]} (\theta)\mathrm{d}\theta,\\
\nonumber&  = \mathbf{1}_{\{t > T_n \}}\mathrm{d}t  \delta_{\mathcal J_n}(\mathrm{d}k, \mathrm{d}e) \mathbf{1}_{[0,\bar \lambda^e]}(\theta)\mathrm{d}\theta .
\end{aligned}
$$ {#eq-barmun}
By definition, $\bar Q^n$ has no jumps before $T_n$.\
As for the first event, the triplet $(T_{n+1}, E_{n+1}, I_{n+1})$ is
determined by the first jump of the measure
$Q^n (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e) := \int_{\mathbb R^+} \mathbf{1}_{\{\theta \leq \lambda_s^e(I_k, Z_{T_n})\}}\bar Q^n (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e, \mathrm{d}\theta)$,
obtained by thinning of $\bar Q^n$. Finally, since the population
composition is constant on $[T_n, T_{n+1}[$, $Z_t = Z_{T_n}$, the
population on $[0,T_{n+1}]$ is defined by: 
$$
\begin{aligned}
\nonumber Z_{t\wedge T_{n+1}}  &  = Z_{t \wedge T_n }  + \mathbf{1}_{\{t\geq T_{n+1}\}}\phi^{E_{n+1}}(T_{n+1}, I_{n+1}), \\
& = Z_{t \wedge T_n } + \int_{t \wedge T_n}^{t \wedge T_{n+1}} \int_{\mathcal{J}_n
\times \mathbb R^+} \phi^e (s , I_k)  \mathbf{1}_{\{\theta\leq \lambda_s^e(I_k, Z_{s^-})\}}\bar Q^n (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e, \mathrm{d}\theta ).
\end{aligned}
$$  {#eq-z-tn} 
Applying $n$ times @eq-z-tn yields
that: 
$$\begin{aligned}
Z_{t\wedge T_{n+1}} = Z_0  + \sum_{l=0}^n \int_{t \wedge T_l}^{t \wedge T_{l+1}}\int_{ \mathcal{J}_l \times  \mathbb R^+} \phi^e (s , I_k)  \mathbf{1}_{\{\theta\leq \lambda_s^e(I_k, \tilde Z_{s^-})\}}\bar Q^l (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e, \mathrm{d}\theta ).
\end{aligned}
$$

**Step 3** Finally, let $\tilde{Z}$ be the solution of @eq-sde-eq-Z,
with $(\tilde T_n)_{n\geq 0}$ the sequence of its event times. Then, we
can write similarly for all $n\geq 0$: $$\begin{aligned}
 \tilde Z_{t\wedge \tilde T_{n+1}} & = Z_0  + \sum_{l=0}^n \int_{t \wedge \tilde  T_l}^{t \wedge \tilde T_{l+1}} \int_{\mathcal{J}\times \mathbb R^+} \phi^e (s , I_k)  \mathbf{1}_{\{\theta\leq \lambda_s^e(I_k,  \tilde Z_{s^-})\}}\mathbf{1}_{ \{k \leq \tilde N_{s^-} \}}   Q(\mathrm{d}s, \mathrm{d}k , \mathrm{d}e , \mathrm{d}\theta ), \\
& = Z_0  + \sum_{l=0}^n \int_{t \wedge \tilde  T_l}^{t \wedge \tilde T_{l+1}} \int_{\mathcal{J}\times \mathbb R^+} \phi^e (s , I_k)  \mathbf{1}_{\{\theta\leq \lambda_s^e(I_k,  \tilde Z_{s^-})\}}\mathbf{1}_{ \{k \leq \tilde N_{\tilde T_l} \}}   Q(\mathrm{d}s, \mathrm{d}k , \mathrm{d}e , \mathrm{d}\theta ),
\end{aligned}$$ since $\tilde  N_{s^-} = \tilde N_{T_l}$ on
$\tilde [T_l, \tilde T_{l+1}]$.\
For each $l \geq 0$, let
$$\tilde Q^l(\mathrm{d}t, \mathrm{d}k , \mathrm{d}e , \mathrm{d}\theta )  = \mathbf{1}_{\{t >  \tilde T_l\}} \mathbf{1}_{ \{1, \dots , \tilde N_{\tilde T_l} \}}(k)   Q(\mathrm{d}t, \mathrm{d}k , \mathrm{d}e , \mathrm{d}\theta ).$$
By @prp-restriction-poisson-measure, $\tilde Q^l$ is,
conditionally to $\mathcal{F}_{T_l}$, a Poisson measure of intensity
$$\mathbf{1}_{\{t >  \tilde T_l\}}   \mathrm{d}t \mathbf{1}_{ \{1, \dots , \tilde N_{\tilde T_l} \}}(k) \delta_{\mathcal{J}}(\mathrm{d}k, \mathrm{d}e) \mathrm{d}\theta.$$
Noticing that
$\mathbf{1}_{ \{1, \dots , \tilde N_{\tilde T_l} \}}(k) \delta_{\mathcal{J}}(\mathrm{d}k, \mathrm{d}e)  =\delta_{\mathcal{J}_l}( \mathrm{d}k , \mathrm{d}e)$,
this shows that $\tilde{Q}^l$ has the conditional intensity $\bar \mu^l$
defined in @eq-barmun and has thus the same distribution than
$\bar Q^l$. Thus, $Z$ in an exact simulation of @eq-sde-eq-Z.
